---
title: 'l1000: tutorial'
author:
    - Bernardo P. de Almeida
    - Nuno Saraiva-Agostinho
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
    \usepackage[utf8]{inputenc}
    %\VignetteIndexEntry{Case study: command-line interface (CLI)}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

---

# Introduction

`l1000` is an R package designed to compare differential gene
expression results with those from known cellular perturbations (such as gene 
knock-down, overexpression or small molecules) derived from the 
[Connectivity Map][clue.io] (Subramanian et al., Cell 2017). Such analyses allow
not only to infer the molecular causes of the observed difference in gene
expression but also to identify small molecules that could drive or revert
specific transcriptomic alterations.

To illustrate the package functionalities, we will use an example based on a 
gene knock-down dataset from the [ENCODE project][ENCODE] for which there is 
available RNA-seq data. After performing differential expression analyses to the 
macthed-control sample, we will compare the respective transcriptomic changes 
with the ones caused by all Connectivity Map's gene knock-down perturbations to
identify which ones have similar or inverse transcriptomic changes to the 
observed ones. As a positive control, we expect to find the knock-down of the 
gene depleted in the ENCODE experiment as one of the most similar transcriptomic
perturbations.

# Getting started

To load the `l1000` package into your R envirnoment type:

```{r eval=FALSE}
library(l1000)
```

# Load ENCODE RNA-seq data and perform differential gene expression analysis

In this example, we will use the EIF4G1 shRNA knockdown followed by RNA-seq experiment in HepG2 cell line from the ENCODE project as the dataset of 
interest. The RNA-seq processed data (gene quantifications from RSEM method) for
the EIF4G1 knock-down and respective controls (two replicates each) can be
downloaded by typing:

```{r}
gene <- "EIF4G1"
cellLine <- "HepG2"

ENCODEfile  <- "~/Documents/L1000_drug_signature/L1000_Rpackage/ENCODE/ENCODE_knockdown_controls.tsv"

ENCODEmetadata <- loadENCODEmetadata(ENCODEfile, cellLine=cellLine, gene=gene,
                                     lab="ENCODE Processing Pipeline",
                                     outputType="gene quantifications")
table(ENCODEmetadata$`Experiment target`)
length(unique(ENCODEmetadata$`Experiment target`))

counts <- loadENCODEgeneExpression(ENCODEmetadata)
```

Gene expression data (read counts) were quantile-normalized using [`voom`][voom] 
and differential expression analysis was performed using the [`limma`][limma] R
package.

```{r}
# Remove low coverage (at least 10 counts shared across two samples)
minReads   <- 10
minSamples <- 2
filter <- rowSums(counts[ , 3:6] >= minReads) >= minSamples
counts <- counts[filter, ]

# Load gene annotation file
geneAnnot <- data.table::fread(
    "/genedata/Resources/Annotations/GeneAnnotation.Human.Ensembl81.txt",
    na.strings=c("NA", "na", "n.a.", ""))

# Perform differential gene expression analysis
diffExpr <- performDifferentialExpression(counts, geneAnnot)
```

As our metric of differential expression after EIF4G1 shRNA knock-down, we will 
use the respective t-statistic.

```{r}
# Get t-statistics of differential expression with respective gene names
diffExprStat <- diffExpr$t
names(diffExprStat) <- diffExpr$Gene_symbol
```

# Load Connectivity Map perturbation data

We will compare our differential gene expression metric with all Connectivity
Map's gene knock-down perturbations in the same cell line (HepG2). Note that 
this comparison can also be done to perturbations in a different cell line or in
all cell lines, using in the last case the average result from comparisons wih 
all cell lines. Differential gene expression for each Connectivity Map's
perturbations are available in z-score normalised values (see Subramanian et 
al., Cell 2017 for more details).

```{r}
# Check available conditions for L1000 perturbations
getL1000Conditions("L1000metadata.txt")

# Code for loading CMap gene KD HepG2 data
l1000perturbations <- loadL1000perturbations(
    "L1000metadata.txt", "L1000zscores.gctx", "L1000geneInfo.txt",
    cellLine="HepG2",
    perturbationType="Consensus signature from shRNAs targeting the same gene")
```

If the interest would be in small molecules, one could download the differential
gene expression z-scores for each small molecule perturbation in HepG2 (given a
specific concentration and time point of RNA extraction) through the following
command:

```{r}
l1000perturbations <- loadL1000perturbations(
    "L1000metadata.txt", "L1000zscores.gctx", "L1000geneInfo.txt",
    cellLine="HepG2", timepoint="96 h", dosage="60 µM",
    perturbationType="Compound")
```

# Comparison with Connectivity Map's perturbations

This is the main function of the package. Here we compare the statistic for
differential expression (the t-statistic, in this case) with the z-scores for 
the perturbations we have loaded using the three different methods available
(Spearman's correlation, Pearson's correlation and Gene Set Enrichment Analysis
(GSEA)). For GSEA, the default option is to use the top and bottom 150 genes
(ranked by the user's t-statistic) as gene sets, but this can be changed by the
user.

(code for comparing with gene KD perturbations using the three methods, just to teach people how to use it, if you agree)

```{r}

```

In case of small molecules:

(code for comparing with small molecules perturbations using the three methods, just to teach people how to use the different methods, if you agree)

For each method used, this function will return a table with the results of the
comparison with each perturbation tested, including the test statistics 
(depending on the method used: Spearman's correlation coefficient, Pearson's
correlation coefficient or GSEA Enrichment Score), the respective p-value and 
the Benjamini-Hochberg-adjusted p-value.

(code for head of output gene KD table for all methods)

# Plot

To analyse the relationship between the user's differential expression with the
one caused by a specific perturbation, we provide the option to create a scatter
plot (For Spearman or Pearson analyses) or a GSEA plot.

Below you can see how to plot the relationship between the ENCODE's EIF4G1 shRNA
knockdown with the perturbations with the most similar or inverse associations.

(code for the three (spearman, pearson and GSEA) plots for top positive (I think
it is the EIF4G1) and top negative gene association)

# Feedback

All feedback on the program, documentation and associated material (including
this tutorial) is welcome. Please send any suggestions and comments to:

> Bernardo P. de Almeida (bernardo.almeida@imp.ac.at)
>
> Nuno Saraiva-Agostinho (nunoagostinho@medicina.ulisboa.pt)
>
> [Disease Transcriptomics Lab, Instituto de Medicina Molecular (Portugal)][iMM]

# References

[iMM]: http://imm.medicina.ulisboa.pt/group/distrans/
[ENCODE]: https://www.encodeproject.org/
[clue.io]: https://clue.io/
[limma]: https://www.ncbi.nlm.nih.gov/pubmed/25605792
[voom]: https://www.ncbi.nlm.nih.gov/pubmed/24485249
